Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-15 Ben_Manuscripts/hdphmm/Draft.tex
Added a few comments
Updates to intro

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables. If one could apply molecular dynamics 
  (MD) simulations to explain important experimental metrics such as solute flux and 
  selectivity in terms of detailed chemically dependent solute motion, then we can more
  easily bridge the disconnect between theory and application. In many systems, it's 
  possible to achieve this by calculating diffusion constants. Unfortunately, for 
  systems which exhibit complex short timescale subdiffusive dynamics, one cannot 
  calculate reliable diffusion constants since solute mean square displacements (MSDs)
  are non-linear on the timescales accessible to molecular simulation. Therefore, we
  are in need of a more flexible approach for modeling solute dynamics which makes 
  less assumptions about the solute's average behavior.

  In this work, we are particularly interested in the design of lyotropic liquid crystal (LLC)
  monomers, a class of amphiphilic molecules whose ordered self-assembled phases can be 
  cross-linked into mechanically strong membranes capable of highly selective separations.
  \begin{itemize}
  	\item Inverted hexagonal, H\textsubscript{II}, phase LLC membranes are characterized
  	by hexagonally packed, uniform-sized and straight pores, an ideal geometry for high 
  	throughput transport.
	\item Their pores are lined with the LLC monomer functional groups which can
	potentially be designed to interact with solutes in a chemically-specific manner.
	\item It would be extremely useful to experimentalists if we could connect LLC monomer
	design with measurable quantities.
	\item This goal is complicated by the non-Brownian hopping and trapping behavior induced
	by the membrane's diverse topology and inhomogeneous structure.~\cite{coscia_understanding_2019,coscia_chemically_2019}
  \end{itemize}

  We have already progressed our understanding of LLC monomer design in our previous work
  by building stochastic time series models based on the chemical intuition gained
  from qualitative MD studies of solute transport in an H\textsubscript{II} phase LLC
  membrane.~\cite{coscia_chemically_2019,coscia_capturing_2020}
  \begin{itemize}
    \item We identified three classes of trapping behavior which result in short time scale
    subdiffusion.
    \item In our first approach to stochastically model this behavior, we treated the
    system's dynamics as a sequence of anti-correlated hops between power law distributed
    periods of entrapment, a framework called subordinated fractional Brownian or L\'evy
    motion.
    \item In a second approach, we treated solute motion as a Markov state model with 
    state-dependent dynamics, where we parameterized the state transition probabilities 
    between each of eight discrete states, defined by the observed trapping mechanism, 
    as well as the solute dynamics within each of these states. 
    \item We showed how one could use realizations of any stochastic model like these in
    order to predict macroscopic flux and selectivity. 
  \end{itemize}

  Although powerful and insightful, our previous work required considerable human effort
  in order to analyze the MD trajectories and to hypothesize stochastic models which
  matched the behavior of solutes.
  \begin{itemize}
    \item It would be significantly more useful and broadly applicable if we could 
    design an approach which automatically distinguishes and parameterizes varied solute 
    dynamics and leaves a facile way to generate an ensemble of stochastic trajectory
    realizations that can be used to predict macroscopic transport properties.
  \end{itemize}

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    %MRS: start with where it has been used before (to establish utility) a
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.~\cite{calderon_inferring_2015,fox_sticky_2011}
    %MRS: Then say it hasn't been used in this context (do we know of any, then cite, if none we have found, say so.)
    \item The IHMM has not been used much for MD trajectory analysis.
  \end{itemize}

@@

  \subsection{Predicting Mean Squared Displacement}\label{method:realizations}

  \textit{Generating Stochastic IHMM Trajectory Realizations}: We used two 
  different methods to stochastically generate trajectories:
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
    \item For each independent MD trajectory, we generated stochastic trajectory
    realizations using the finalized parameters of the associated IHMM. This 
    results in 24 separate ensembles of trajectories for each solute studied. 
    \item Using the clustered parameter sets, we generated a single ensemble
	of trajectories. Realizations are representative of the combined behavior
	of all solute trajectories.
  \end{enumerate}

@@

  \subsection{Predicting Selectivity}\label{method:selectivity}

@@

  %BJC4: It might be good to add a panel to this figure which gives a radial view of the pore
  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{rz_unclustered_MET.pdf}
  \caption{The IHMM found 24 distinct VAR(1) states in this representative methanol trajectory.
  We performed the parameterization in 3D cartesian coordinates, but for this figure, we 
  converted to cylindrical coordinates, with $r$ the distance from the pore center, to 
  provide a clearer picture of solute motion.}\label{fig:rz_unclustered}
  \end{figure}

@@

%BJC4: the following has pretty all been replaced.
%  Most of what we 
%  have learned thus far has been based on the parameters of our model with some 
%  qualitative validation using a single methanol trajectory. While 
%  we may be able to further corroborate our arguments by analyzing more
%  individual trajectories, it is more useful to correlate the state behavior with
%  solute-membrane interactions.
%
%  %BJC2: A lot of the following will probably be reworked with different clustering scheme
%%  \textit{Hydrogen Bonding Patterns}: 
%  The types of hydrogen bonding undergone by solutes is dependent on radial
%  location (see Figure~\ref{fig:hbond_pichart}). States with radial means near 1 nm
%  or less (states 1, 3 and 5) typically hydrogen bond with the monomer carboxylate
%  groups and ether linkages. States with radial means greater than 1 nm (states 2,
%  4, 6 and 7) tend to hydrogen bond with the monomer tails. Our previous work has
%  shown that there is a gradual transition from the hydrophilic to hydrophobic region in
%  these membranes, meaning the monomer location is not tightly bound in $r$. 
%  %MRS: above logic is not clear?
%  Therefore, there are instances of hydrogen bond interactions that one might consider 
%  uncharacteristic based on the state's radial mean.
%  %MRS: what is implication of ``uncharacteristicness''? 
%
%  States which exhibit longer dwell times have longer hydrogen bond lifetimes (see
%  Figure~\ref{fig:hbond_pichart}). States 3--7 have significantly longer hydrogen 
%  bond lifetimes than states 1 and 2. It makes physical and mathematical sense that
%  states 1 and 2 should not stay trapped by hydrogen bonding for long periods because
%  they typically take large hops as communicated by their large covariances.
%  In the tail states, hydrogen bond lifetimes appear to increase with their dwell 
%  times. However, in the pores, the same linear relationship does not hold.
%  For example, state 5 has the lowest hydrogen bond lifetime of states 3--7
%  despite having the second longest dwell times.

%  The local density experienced by solutes may have an impact on the length of solute 
%  entrapment. On average, solutes in state 5 are surrounded by a relatively high density
%  of particles. Based on its radial location and preference towards hydrogen bonding
%  with the ether linkages (see Figure~\ref{fig:hbond_pi_charts}), it is reasonable to
%  assume that while in state 5, methanol spends most of its time just outside the 
%  pore region, sandwiched between the monomer head groups. Although state 5 hydrogen
%  bonds break more frequently than one might expect for a state with long dwell times,
%  it is possible that the solute's dense surroundings prevent it from wandering off 
%  before the hydrogen bonding interaction can be recovered. This may explain the 
%  behavior of the methanol in state 5 of the trajectory in Figure~\ref{fig:mechanism_map}
%  where the solute's center of mass appears to move more freely about its mean than 
%  other trapped states. The data suggests that methanol molecules in state 3 lie closer
%  to the pore center than state 5 because they tend to hydrogen bond most frequently 
%  with the carboxylate groups. Methanol molecules which experience state 3 are subject 
%  to a lower density local environment and rely more heavily on hydrogen bonding interactions, 
%  evidenced by their longer lifetimes, to stay trapped.
%  
%  States describing behavior in the tail region show a direct relationship between 
%  hydrogen bond lifetimes and local density. Methanol molecules which experience state 2 
%  spend their time in a low density region of the tails where they can easily break free
%  of hydrogen bonds. Conversely, methanol molecules trapped in state 7 are in the densest 
%  region of all states studied. It is likely difficult for methanol to escape these high 
%  density regions and its position is further stabilized via hydrogen bonding.
%  
%  %BJC: I didn't put sodium ion association in this figure because it's infrequent. 
%  %BJC: If I compare to acetic acid or urea, I could add it back. I was thinking of splitting
%  % the pi chart in half with slices corresponding to hbonds on one side and association on the other
%  \begin{figure}
%  \centering
%  \begin{subfigure}{0.54\textwidth}
%  \includegraphics[width=\textwidth]{hbond_pi_charts.pdf}
%  \caption{}\label{fig:hbond_pi_charts}
%  \end{subfigure}
%  \begin{subfigure}{0.45\textwidth}
%  \includegraphics[width=\textwidth]{hbond_lifetimes_density_MET.pdf}
%  \caption{}\label{fig:hbond_lifetimes_density_MET}
%  \end{subfigure}
%  %MRS: legend doesn't show stripes.
%  \caption{(a) Methanol hydrogen bonds in different segments of the LLC monomer
%  dependent on the state. States with mean radial locations in the pore region tend to
%  hydrogen bond with the head groups and ether linkages. States with mean radial
%  locations in the tails tend to hydrogen bond with the tail oxygen atoms.
%  (b) Hydrogen bond lifetimes and the local density experienced by methanol
%  molecules can help us to understand dynamical states in terms of membrane 
%  interactions. The bars are colored according to match the states in Figures~\ref{fig:mechanism_map}
%  and~\ref{fig:common_states_MET}. States 1 and 2, which have large covariances, also have short
%  hydrogen bond lifetimes and exist in relatively low density regions. The
%  hydrogen bond lifetimes of methanol in states with radial means in the tail
%  region tend to increase with local density and mean dwell times. 
%  %MRS7: not clear from graphic which ones are in the pore.
%  %MRS7: is there any legend for the color for b? If same as above figure, indicate. Don't REALLY need color since they bars are already labeled, so it's 2 sources of duplicate information.
%  %BJC7: The state numbers are all you need to compare to the previous figure BUT the color make it easier 
%  %plot of the methanol trajectory where states aren't labeled.
%  %MRS7: cross-hatching doesn't appear in the legend? Not sure which is which (I assume hatched is number density because closer to that axis).
%  %BJC7: Is it because the hatches aren't dense enough? I see them in the legend
%  This monotonic
%  relationship does not exist for states in the pore region. The high 
%  local density experienced by state 5 methanol molecules may explain why 
%  hydrogen bond lifetimes are short but dwell times are high. State 5 methanol's
%  preference towards hydrogen bonds with ether linkages suggests that the solute
%  sits between LLC monomer head group which prevent it from wandering to far
%  before recovering its previous hydrogen bond.
%  }\label{fig:hbond_pichart}
%  \end{figure}

>>>>

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables. If one could apply molecular dynamics 
  (MD) simulations to explain important experimental metrics such as solute flux and 
  selectivity in terms of detailed chemically dependent solute motion, then we can more
  easily bridge the disconnect between theory and application. In many systems, it's 
  possible to achieve this by calculating diffusion constants. Unfortunately, for 
  systems which exhibit complex short timescale subdiffusive dynamics, one cannot 
  calculate reliably compute diffusion constants since solute mean square 
  displacements (MSDs) are non-linear on the timescales accessible to molecular simulation.
  Therefore, we are in need of a more flexible approach for modeling solute dynamics
  which makes less assumptions about the solute's short time scale behavior.

  In this work, we are particularly interested in the design of lyotropic liquid 
  crystal (LLC) monomers, a class of amphiphilic molecules whose ordered self-assembled
  phases can be cross-linked into mechanically strong membranes capable of highly 
  selective separations. Inverted hexagonal, H\textsubscript{II}, phase LLC membranes
  are characterized	by hexagonally packed, uniform-sized and straight pores, an ideal
  geometry for high throughput transport. Their pores are lined with the LLC monomer's
  functional groups which can potentially be designed to interact with solutes in a 
  chemically-specific manner. It would be extremely useful to experimentalists if we 
  could connect LLC monomer design with measurable quantities. This goal is complicated
  by the non-Brownian hopping and trapping behavior induced by the membrane's diverse
  topology and inhomogeneous structure.~\cite{coscia_understanding_2019,coscia_chemically_2019}

  We have already progressed our understanding of LLC monomer design in our previous work
  by building stochastic time series models based on the chemical intuition gained
  from qualitative MD studies of solute transport in an H\textsubscript{II} phase LLC
  membrane.~\cite{coscia_chemically_2019,coscia_capturing_2020} We identified three 
  classes of trapping behavior which result in short time scale subdiffusion. In our 
  first approach to stochastically model this behavior, we treated the system's dynamics
  as a sequence of anti-correlated hops between power law distributed periods of 
  entrapment, a framework called subordinated fractional Brownian or L\'evy
  motion.~\cite{thiel_weak_2014,teuerle_modeling_2013} In a second approach, we treated 
  solute motion as a Markov state model~\cite{pande_everything_2010} with state-dependent
  dynamics, where we parameterized the state transition probabilities between each of
  eight discrete states, defined by the observed trapping mechanisms, as well as the 
  solute dynamics within each of these states. We showed how one could use realizations
  of any stochastic model like these in order to predict macroscopic flux and selectivity.

  Although powerful and insightful, our previous work required considerable human effort
  in order to analyze the MD trajectories and to hypothesize stochastic models which
  matched the behavior of solutes. It would be significantly more useful and broadly 
  applicable if we could design an approach which automatically distinguishes and 
  parameterizes varied solute dynamics and leaves a facile way to generate an ensemble
  of stochastic trajectory realizations that can be used to predict macroscopic transport
  properties.

  In this work we attempt to minimize human effort through a novel application of the 
  the sticky hierarchical Dirichlet process autoregressive hidden markov model 
  (HDP-AR-HMM), nonparameteric Bayesian classification technique that allows us to 
  automatically detect and infer the parameters of an unknown number of latent vector
  autoregressive (VAR) modes present in solute center-of-mass time series 
  trajectories.~\cite{fox_bayesian_2010} The sticky HDP-AR-HMM greatly extends the 
  basic framework of more limited finite hidden Markov modeling approaches.

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm is the most commonly used way to determine the hidden
  state sequence and the underlying parameters of a standard HMM.~\cite{baum_maximization_1970}
  It also possible to achieve the same results using a Bayesian 
  framework.~\cite{scott_bayesian_2002,jasra_markov_2005} One can place priors
  on the transition probabilities and parameters of the hidden states' emission
  distributions in order to iteratively infer their values and to estimate the hidden state
  sequence. HMMs have been used in a broad range of applications including speech
  recognition, weather forecasting and bioinformatics.~\cite{juang_on_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}
  %BJC4: comment on usage in MD simulations. MSMs?

  By placing a hierarchical Dirichlet process (HDP) prior on the transition 
  probabilities, it is possible to extend HMMs to an unknown number of hidden
  states.~\cite{teh_hierarchical_2006}
  \begin{itemize}
    \item We will refer to this as the HDP-HMM, although in literature it is 
    often called the infinite hidden Markov model.~\cite{beal_infinite_2002}
    \item The HDP-HMM elegantly overcomes the main shortcoming of standard
    HMMs, the need to choose the number of states, making its use more 
    advantageous in a diversity of applications.
    \item

    diverse applicability and improvement upon the shortcoming 
    of standard HMMs
    \item Applications...speaker diarization, etc.
  \end{itemize}

  One issue with the original formulation of HDP-HMM, is that it does not 
  distinguish self-transitions and transitions between different states.
  \begin{itemize}
    \item Practically, this can result in rapid switching between states and
    inferred state sequence representations that do a poor job of describing
    the observations. 
    \item Fox et al. addressed this issue by adding a `sticky' parameter.~\cite{fox_sticky_2007}
    \item The sticky HDP-HMM is useful in situations where the user has
    prior knowledge that state persistence is likely.
    \item This is important for our work where we expect state persistence 
    due to trapping.
    \item Applications: Speaker diarization, SPT (Chris' diffusion papers)
  \end{itemize}

  The usefulness of the sticky HDP-HMM has been further extended to include 
  temporal dependencies common to many real datasets.
  \begin{itemize}
    \item Fox et al. showed how to incorporate Markov jump linear systems (MJLSs)
    into infinite state models.~\cite{fox_nonparameteric_2009}
    \item An MJLS models each dynamical mode as a linear dynamical process.
    \item There are two MJLS classes: switching linear dynamical systems (SLDSs) 
    and switching vector autoregressive (VAR) processes.
    \item Switching VAR processes are similar to SLDSs except that they do not
    incorporate measurement noise.
    \item Since MD simulations do not contain measurement noise, we will model
    describe the dynamical modes exhibited by solutes in our MD simulations as 
    VAR processes and we will refer to our model as an application of the 
    HDP-AR-HMM, maintaining a distinction from HDP-SLDS models which can
    be useful for experimental measurement where there is measurement noise.
  \end{itemize}

  To our knowledge, this is a novel application of the HDP-AR-HMM.
  \begin{itemize}
    \item We use the temporally-correlated parameters of the model in order
    to learn physical mechanisms and describe solute kinetics.
    \item We go beyond the typical application of IHMMs.
  \end{itemize}

@@

  \subsection{Predicting Mean Squared Displacement}\label{method:realizations}
  %BJC4: reworked this section
  \textit{Generating Stochastic IHMM Trajectory Realizations}: We used two 
  different methods to stochastically generate trajectories:
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
    \item For each independent MD trajectory, we generated stochastic trajectory
    realizations using the finalized parameters of the associated IHMM. This 
    results in 24 separate ensembles of trajectories for each solute studied. 
    \item Using the clustered parameter sets, we generated a single ensemble
	of trajectories. Realizations are representative of the combined behavior
	of all solute trajectories.
  \end{enumerate}

@@

  %BJC4: new section below
  \subsection{Predicting Selectivity}\label{method:selectivity}

@@

  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{rz_unclustered_MET.pdf}
  \caption{The IHMM found 24 distinct VAR(1) states in this representative methanol trajectory.
  We performed the parameterization in 3D cartesian coordinates, but for this figure, we 
  converted to cylindrical coordinates, with $r$ the distance from the pore center, to 
  provide a clearer picture of solute motion.}\label{fig:rz_unclustered}
  \end{figure}

@@
